cardinalR: Generating interesting high-dimensional data structures

### 2025-06-24 paper-cardinalR.Rmd
Updated branching last two sections

Branch $1$ is generated from $X_1 \sim U(0, 1)$, and $X_2 = 0.1 \cdot X_1 + 1 \cdot X_1^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This produces a gently upward-curving parabola in the right half-plane. Branch $2$ is generated from $X_{1} \sim U(-1, 0)$, and $X_2 = 0.1 \cdot X_{1} - 2 \cdot X_{1}^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This creates a steeper, leftward-facing curve in the left half-plane. Branches $3$ to $k$ are added iteratively. Each new branch $i$ begins at a randomly chosen point within a restricted horizontal band: $X_{1} \in [-0.15, 0.15]$ (to ensure connectivity with earlier branches), spans a unit-length interval on $X_{1}$:$X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]$, and has the structure:

@@

The `gen_orglinearbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ approximately linear branches embedded in a $p$-dimensional space. Each branch lies primarily within a distinct $2\text{-}D$ subspace, while the remaining $p - 2$ dimensions contain Gaussian noise.

To construct each branch, a unique or repeated pair of dimensions is selected from the $\binom{p}{2}$ possible $2\text{-}D$ combinations. If $k \leq \binom{p}{2}$, combinations are sampled without replacement. If $k > \binom{p}{2}$, additional pairs are sampled with replacement to reach $k$ total branches. Each selected pair $(x_{i1}, x_{i2})$ defines the $2\text{-}D$ plane for branch $i$.

For branch $i$, $n_i$ points are generated, where $\sum_{i=1}^k n_i = n$. The structure follows:

$$
x_{i1} \sim U(0, 2), \quad
x_{i2} = -s_i \cdot x_{i1} + \epsilon, \quad \epsilon \sim U(0, 0.5),
$$

where $s_i$ is a scale factor controlling the slope. When $k \leq \binom{p}{2}$, $s_i = 1$. When sampling with replacement, $s_i$ is drawn from the set ${1, 1.5, 2, \dots, 8}$ in increments of $0.5$.

For $p > 2$, the remaining dimensions contain independent Gaussian noise:

$$
x_j \sim N(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}.
$$

@@

Let $\binom{p}{2}$ denote the number of unique $2\text{-}D$ subspace combinations. When $k \leq \binom{p}{2}$, $k$ distinct subspace pairs $(x_{i1}, x_{i2})$ are sampled without replacement. Otherwise, combinations are selected with replacement to reach $k$ total branches. For each branch $i = 1, \dots, k$, a random scale factor $s_i$ is used to vary the curvature:

* If $k \leq \binom{p}{2}$: $s_i = 1$
* If $k > \binom{p}{2}$: $s_i \sim \text{Uniform sample from } {1, 1.5, \dots, 8}$

Each branch contains $n_i$ points such that $\sum_{i=1}^k n_i = n$, where the vector $(n_1, \dots, n_k)$ is randomly drawn using the helper function `gen_nsum()` to partition $n$.

Within its assigned subspace $(x_{i1}, x_{i2})$, branch $i$ is defined by:

$$
x_{i1} \sim U(0, 2), \quad
x_{i2} = -s_i \cdot x_{i1}^2 + \epsilon, \quad \epsilon \sim U(0, 0.5)
$$

This forms a smooth downward-opening parabola in the plane defined by $(x_{i1}, x_{i2})$, with the degree of curvature controlled by $s_i$.

For $p > 2$, all remaining dimensions are independent Gaussian noise:

$$
x_j \sim N(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}.
$$

>>>>

Branch $1$ is generated from $X_1 \sim U(0, 1)$, and $X_2 = 0.1X_1 + X_1^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This produces a gently upward-curving parabola in the right half-plane. Branch $2$ is generated from $X_{1} \sim U(-1, 0)$, and $X_2 = 0.1X_{1} - 2X_{1}^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This creates a steeper, leftward-facing curve in the left half-plane. Branches $3$ to $k$ are added iteratively. Each new branch $i$ begins at a randomly chosen point within a restricted horizontal band: $X_{1} \in [-0.15, 0.15]$ (to ensure connectivity with earlier branches), spans a unit-length interval on $X_{1}$:$X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]$, and has the structure:

@@

The `gen_orglinearbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ approximately linear branches embedded in $p\text{-}D$. Each branch lies primarily within a distinct $2\text{-}D$ subspace, while the remaining $p - 2$ dimensions contain Gaussian noise.

To construct each branch, a unique or repeated pair of dimensions is selected from the $\binom{p}{2}$ possible $2\text{-}D$ combinations. If $k \leq \binom{p}{2}$, combinations are sampled without replacement. If $k > \binom{p}{2}$, additional pairs are sampled with replacement to reach $k$ total branches. Each selected pair $(X_{i1}, X_{i2})$ defines the $2\text{-}D$ plane for branch $i$.

For branch $i$, $n_i$ points are generated, where $\sum_{i=1}^k n_i = n$. The structure follows: $X_{i1} \sim U(0, 2)$, and $X_{i2} = -s_i \cdot x_{i1} + \epsilon, \quad \epsilon \sim U(0, 0.5)$, where $s_i$ is a scale factor controlling the slope. When $k \leq \binom{p}{2}$, $s_i = 1$. When sampling with replacement, $s_i$ is drawn from the set ${1, 1.5, 2, \dots, 8}$ in increments of $0.5$. For $p > 2$, the remaining dimensions contain independent Gaussian noise: $X_j \sim N(0, 0.1^2)$, for $j \notin \{i1, i2\}$.

@@

Let $\binom{p}{2}$ denote the number of unique $2\text{-}D$ subspace combinations. When $k \leq \binom{p}{2}$, $k$ distinct subspace pairs $(X_{i1}, X_{i2})$ are sampled without replacement. Otherwise, combinations are selected with replacement to reach $k$ total branches. For each branch $i = 1, \dots, k$, a random scale factor $s_i$ is used to vary the curvature; If $k \leq \binom{p}{2}$: $s_i = 1$, or If $k > \binom{p}{2}$: $s_i \sim \text{Uniform sample from } {1, 1.5, \dots, 8}$.

Each branch contains $n_i$ points such that $\sum_{i=1}^k n_i = n$, where the vector $(n_1, \dots, n_k)$ is randomly drawn using the helper function `gen_nsum()` to partition $n$. Within its assigned subspace $(X_{i1}, X_{i2})$, branch $i$ is defined by: $X_{i1} \sim U(0, 2)$, and $X_{i2} = -s_i \cdot X_{i1}^2 + \epsilon, \quad \epsilon \sim U(0, 0.5)$. This forms a smooth downward-opening parabola in the plane defined by $(X_{i1}, X_{i2})$, with the degree of curvature controlled by $s_i$. For $p > 2$, all remaining dimensions are independent Gaussian noise: $X_j \sim N(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}$.

### 2025-06-24 paper-cardinalR.tex
Updated branching last two sections

Branch \(1\) is generated from \(X_1 \sim U(0, 1)\), and \(X_2 = 0.1 \cdot X_1 + 1 \cdot X_1^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This produces a gently upward-curving parabola in the right half-plane. Branch \(2\) is generated from \(X_{1} \sim U(-1, 0)\), and \(X_2 = 0.1 \cdot X_{1} - 2 \cdot X_{1}^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This creates a steeper, leftward-facing curve in the left half-plane. Branches \(3\) to \(k\) are added iteratively. Each new branch \(i\) begins at a randomly chosen point within a restricted horizontal band: \(X_{1} \in [-0.15, 0.15]\) (to ensure connectivity with earlier branches), spans a unit-length interval on \(X_{1}\):\(X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]\), and has the structure:

@@

The \texttt{gen\_orglinearbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) approximately linear branches embedded in a \(p\)-dimensional space. Each branch lies primarily within a distinct \(2\text{-}D\) subspace, while the remaining \(p - 2\) dimensions contain Gaussian noise.

To construct each branch, a unique or repeated pair of dimensions is selected from the \(\binom{p}{2}\) possible \(2\text{-}D\) combinations. If \(k \leq \binom{p}{2}\), combinations are sampled without replacement. If \(k > \binom{p}{2}\), additional pairs are sampled with replacement to reach \(k\) total branches. Each selected pair \((x_{i1}, x_{i2})\) defines the \(2\text{-}D\) plane for branch \(i\).

For branch \(i\), \(n_i\) points are generated, where \(\sum_{i=1}^k n_i = n\). The structure follows:

\[
x_{i1} \sim U(0, 2), \quad
x_{i2} = -s_i \cdot x_{i1} + \epsilon, \quad \epsilon \sim U(0, 0.5),
\]

where \(s_i\) is a scale factor controlling the slope. When \(k \leq \binom{p}{2}\), \(s_i = 1\). When sampling with replacement, \(s_i\) is drawn from the set \({1, 1.5, 2, \dots, 8}\) in increments of \(0.5\).

For \(p > 2\), the remaining dimensions contain independent Gaussian noise:

\[
x_j \sim N(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}.
\]

@@

Let \(\binom{p}{2}\) denote the number of unique \(2\text{-}D\) subspace combinations. When \(k \leq \binom{p}{2}\), \(k\) distinct subspace pairs \((x_{i1}, x_{i2})\) are sampled without replacement. Otherwise, combinations are selected with replacement to reach \(k\) total branches. For each branch \(i = 1, \dots, k\), a random scale factor \(s_i\) is used to vary the curvature:

\begin{itemize}
\tightlist
\item
  If \(k \leq \binom{p}{2}\): \(s_i = 1\)
\item
  If \(k > \binom{p}{2}\): \(s_i \sim \text{Uniform sample from } {1, 1.5, \dots, 8}\)
\end{itemize}

Each branch contains \(n_i\) points such that \(\sum_{i=1}^k n_i = n\), where the vector \((n_1, \dots, n_k)\) is randomly drawn using the helper function \texttt{gen\_nsum()} to partition \(n\).

Within its assigned subspace \((x_{i1}, x_{i2})\), branch \(i\) is defined by:

\[
x_{i1} \sim U(0, 2), \quad
x_{i2} = -s_i \cdot x_{i1}^2 + \epsilon, \quad \epsilon \sim U(0, 0.5)
\]

This forms a smooth downward-opening parabola in the plane defined by \((x_{i1}, x_{i2})\), with the degree of curvature controlled by \(s_i\).

For \(p > 2\), all remaining dimensions are independent Gaussian noise:

\[
x_j \sim N(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}.
\]

>>>>

Branch \(1\) is generated from \(X_1 \sim U(0, 1)\), and \(X_2 = 0.1X_1 + X_1^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This produces a gently upward-curving parabola in the right half-plane. Branch \(2\) is generated from \(X_{1} \sim U(-1, 0)\), and \(X_2 = 0.1X_{1} - 2X_{1}^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This creates a steeper, leftward-facing curve in the left half-plane. Branches \(3\) to \(k\) are added iteratively. Each new branch \(i\) begins at a randomly chosen point within a restricted horizontal band: \(X_{1} \in [-0.15, 0.15]\) (to ensure connectivity with earlier branches), spans a unit-length interval on \(X_{1}\):\(X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]\), and has the structure:

@@

The \texttt{gen\_orglinearbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) approximately linear branches embedded in \(p\text{-}D\). Each branch lies primarily within a distinct \(2\text{-}D\) subspace, while the remaining \(p - 2\) dimensions contain Gaussian noise.

To construct each branch, a unique or repeated pair of dimensions is selected from the \(\binom{p}{2}\) possible \(2\text{-}D\) combinations. If \(k \leq \binom{p}{2}\), combinations are sampled without replacement. If \(k > \binom{p}{2}\), additional pairs are sampled with replacement to reach \(k\) total branches. Each selected pair \((X_{i1}, X_{i2})\) defines the \(2\text{-}D\) plane for branch \(i\).

For branch \(i\), \(n_i\) points are generated, where \(\sum_{i=1}^k n_i = n\). The structure follows: \(X_{i1} \sim U(0, 2)\), and \(X_{i2} = -s_i \cdot x_{i1} + \epsilon, \quad \epsilon \sim U(0, 0.5)\), where \(s_i\) is a scale factor controlling the slope. When \(k \leq \binom{p}{2}\), \(s_i = 1\). When sampling with replacement, \(s_i\) is drawn from the set \({1, 1.5, 2, \dots, 8}\) in increments of \(0.5\). For \(p > 2\), the remaining dimensions contain independent Gaussian noise: \(X_j \sim N(0, 0.1^2)\), for \(j \notin \{i1, i2\}\).

@@

Let \(\binom{p}{2}\) denote the number of unique \(2\text{-}D\) subspace combinations. When \(k \leq \binom{p}{2}\), \(k\) distinct subspace pairs \((X_{i1}, X_{i2})\) are sampled without replacement. Otherwise, combinations are selected with replacement to reach \(k\) total branches. For each branch \(i = 1, \dots, k\), a random scale factor \(s_i\) is used to vary the curvature; If \(k \leq \binom{p}{2}\): \(s_i = 1\), or If \(k > \binom{p}{2}\): \(s_i \sim \text{Uniform sample from } {1, 1.5, \dots, 8}\).

Each branch contains \(n_i\) points such that \(\sum_{i=1}^k n_i = n\), where the vector \((n_1, \dots, n_k)\) is randomly drawn using the helper function \texttt{gen\_nsum()} to partition \(n\). Within its assigned subspace \((X_{i1}, X_{i2})\), branch \(i\) is defined by: \(X_{i1} \sim U(0, 2)\), and \(X_{i2} = -s_i \cdot X_{i1}^2 + \epsilon, \quad \epsilon \sim U(0, 0.5)\). This forms a smooth downward-opening parabola in the plane defined by \((X_{i1}, X_{i2})\), with the degree of curvature controlled by \(s_i\). For \(p > 2\), all remaining dimensions are independent Gaussian noise: \(X_j \sim N(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}\).
